---
title: "Northwestern R Users Group: Network Analysis in R"
author: "Richard Paquin Morel"
date: "April 5, 2018"
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
```

### What is network analysis?

- Relational analysis
- The unit of analysis is, at minimum, a _dyad_, some relationship among some entities
- Relevant to a variety of fields -- social sciences, engineering, computer science, biological and physical sciences, mathematics, etc.

### Special challenges for network analysis

- Data format and storage is complex
- Computationally expensive
- Standard assumptions of statistical analysis often do not hold
  - Problems of, e.g., autocorrelation

#### Data format and storage is complex

- Object representing relationships among entities (e.g., advice seeking among colleagues in a school)
- Object representing attributes of entities (e.g,. characteristics of those colleagues)

#### Example
```{r example network, echo = T, message = F, warning = F}
# install.packages(c("igraph", "statnet", "ggraph"))

library(tidyverse)
library(ggraph)
data(faux.desert.high, package = "ergm")
ggraph(faux.desert.high) +
  geom_node_point(aes(color = as.factor(grade)), alpha = 1) +
  geom_edge_link(alpha = 0.2) +
  guides(color = guide_legend("Grade")) +
  theme_void()
```

#### Computationally expensive

A directed network with...

- 3 nodes $\rightarrow$ $2^{3(3-1)} = 64$ possible networks
- 10 nodes $\rightarrow$ $2^{10(10-1)} = 1.2379 * 10^{27}$ possible networks
- 25 nodes $\rightarrow$ $2^{600} >$ the estimated number of atoms in the universe


#### Violation of standard statistical assumptions

- Observations in networks are inherently interdependent
- Probability of ties is not unconditional
- The formation of a tie in a network impacts the probability of the formation of ties throughout the entire network

### Housekeeping 

#### Some vocabulary

| Thing         | Graph Theory Term | Social Network Term|
|---------------|-------------------|--------------------|
| Entity        | Vertex            | Node, actor        |
| Relationship  | Edge              | Tie                |
| Set of entities and relationships | Graph | Network    |
| Nature of relationship | Graph/digraph | Undirected/directed |

- If the relationship has some inherent directionality (e.g., "goes to for advice"), it is called a _directed_ network, or digraph. If not it is an _undirected_ network, or graph

#### What I will talk about

- The nitty gritty of network analysis in R
  - The high points and low points of the two big network packages - igraph and the statnet suite of packages
  - Creating data
  - Manipulating data
  - Key functions from these packages
  - Visualization
  
### Network analysis in R: igraph vs. statnet

#### `igraph`
- Data handling/manipulation/visualization
- Comprehensive set of functions for graph theory and network analysis
- Python and C editions
- Extensive support on Stack Overflow

#### `statnet`
  
- A suite of packages
  - Data handling/manipulation/visualization --> `network`
  - Standard tools for network analysis --> `sna`
  - Tools for statistical inference --> `ergm` (and extensions), `latentnet`
  - Etc. --> `EpiModel` for epidemic modeling

#### Which to use?
- Probably both
  - `statnet` --> social networks and statistical models of networks
  - `igraph` --> graph theory and mathematical models of graphs
- `statnet` created and maintained by statisticians and sociologists
- `igraph` created and maintained by computer scientists

### How are networks represented?

- Several formats -- adjacency matrix, edgelist, incidence matrix

#### The adjacency matrix

- An _n_ by _n_ matrix where rows/columns represent entities and each _i, j_ cell represents the present or absence of a relationship between _i_ and _j_.
```{r load packages and data, message = F, echo = T}

### load packages I will use statnet for now.

library(tidyverse)
library(statnet)

data("faux.desert.high") # a fake dataset of friendships in a high school simulated from real data
net <- faux.desert.high # giving it an easier name
```

```{r adjacency matrix}
as.matrix(net)[7:20, 7:20]
```

#### The edgelist

- A two-column data frame with a "To" and "From" column where each row represent a relationship between two entities.

```{r edgelist}
as.edgelist(net)[1:10, ]
```

### Creating and manipulating network data

#### Prepping data...

```{r prepping data for Creating, echo = T}
data(faux.desert.high)
adj_matrix <- as.matrix(faux.desert.high)
el <- as.edgelist(faux.desert.high) %>% 
  as.data.frame()
colnames(el) <- c("to", "from")
dhs_attr <- data_frame(id = paste("DHS", net %v% "vertex.names", sep = "_"),
                        grade = net %v% "grade",
                        race = net %v% "race",
                        sex = net %v% "sex")
colnames(adj_matrix) <- dhs_attr$id
rownames(adj_matrix) <- dhs_attr$id
el <- 
  el %>% 
  transmute(to = dhs_attr$id[match(el$to, rownames(dhs_attr))],
            from = dhs_attr$id[match(el$from, rownames(dhs_attr))])
```

- Three objects:
  - `adj_matrix` - an adjacency matrix representing friendships among students at Desert High
  - `el` - an edgelist with the same information
  - `dhs_attr` - a data frame of student attributes 
  
```{r data to Create}
glimpse(dhs_attr)
```

#### `statnet` approach

- Create a network object: `network` or `as.network`


```{r}
### network(x, 
###         vertex.attr=NULL, 
###         vertex.attrnames=NULL, 
###         directed=TRUE, 
###         hyper=FALSE, loops=FALSE, multiple=FALSE, bipartite = FALSE, ...)
```

- `x` can be either an adjacency matrix or an edgelist

#### `statnet` approach

```{r statnet Create}
library(statnet)

net <- network(adj_matrix)
net
```

#### `statnet` approach
```{r Create with attributes}
network(adj_matrix, 
        vertex.attr = list(dhs_attr$grade,
                           dhs_attr$race,
                           dhs_attr$sex),
        vertex.attrnames = list("grade", 
                                "race",
                                "sex"))
```

#### `igraph` approach

- `graph_from_*` syntax
- For adjacency matrices: `graph_from_adjacency_matrix`
- For edgelists: `graph_from_edgelist` or `graph_from_data_frame`

```{r igraph Create, message = F, warning = F}
library(igraph)

g <- graph_from_adjacency_matrix(adj_matrix)
g
```

### A word of caution, re: edgelists

#### Edgelists
- Edgelists do not have information for entities with no relationships (isolates)

```{r edgelists}
graph_from_data_frame(el)
```

#### Add the missing nodes

```{r add missing nodes}
el_net <- graph_from_data_frame(el)
dhs_attr$id[which(!dhs_attr$id %in% V(el_net)$name)]
el_net <- igraph::add_vertices(el_net, 7,
                     name = dhs_attr$id[which(!dhs_attr$id %in% V(el_net)$name)])
```

### Manipulating networks

- Each package has its own special syntax for interacting with networks

#### Accessing and manipulating attributes

|package    | class     | vertices | edges | networks |
|-----------|-----------|----------|-------|----------|
| `statnet` | "network" | `%v%`    | `%e%` | `%n%`    |
| `igraph`  | "igraph"  | `V()`    | `E()` | `$`      |


#### In `statnet`
```{r manipulating network objects}
net <- network(adj_matrix)
network::list.vertex.attributes(net)

net %v% "race" <- dhs_attr$race
net %v% "grade" <- dhs_attr$grade
net %v% "grade"
```

#### In `igraph`...

```{r manipulating igraph objects}
vertex_attr_names(g)

V(g)$race <- dhs_attr$race
V(g)$grade <- dhs_attr$grade
V(g)$grade
```

#### Subsetting networks

- Various reasons to subset a network 
  - Sub-network of friendships among 8th graders
  - Sub-network of friendships for informal subgroups
- `igraph::induced_subgraph(network, vertices)`
- `statnet` - `network %s% vertices`

#### Subsetting

```{r subsetting}
induced_subgraph(g, which(V(g)$grade == 8))

### net %s% which((net %v% "grade") == 8)
### if `igraph` is loaded:
### network::`%s%`(net, which((net %v% "grade") == 8))
```


### Sampling random networks

#### `igraph` is the real winner here

- `igraph` has many (>30) functions for producing random graphs
- Syntax - `sample_*`
- Scale free, uniform probability, small world, Erdös-Rényi, preferential attachment, etc., etc.

#### A scale-free network
```{r}
samp_g <- sample_pa(100)
qplot(igraph::degree(samp_g), geom = "density")
```

#### An Erdös-Rényi or Bernoulli network
```{r}
samp_g <- sample_gnp(100, 0.25)
qplot(igraph::degree(samp_g), geom = "density")
```

#### `statnet` has options...

- Just not as many or as robust
- `rgraph` --> Bernoulli
- `rgnm` --> conditioned on density
- `rguman` --> conditioned on dyad census

### Key network measures

#### Both packages have all standard network measures

- Node-level indices - Properties of vertices
  - Centrality
- Graph-level indices - Properties of networks
  - Density, centralization, dyad and triad censuses
- Subgraph and community detection
  - Blockmodels, cliques, components, communities
  
#### Node-level indicies

- Centrality captures the prominence, relative importance, influence of a vertex
- Most basic is degree centrality -- how many edges does a vertex have? How active is a person? How popular?
  - Is the network directed or undirected?

```{r centrality}
### sna::degree(net, cmode = "freeman")
igraph::degree(g, mode = "total")
```

#### Node level indices, continued

- Betweenness centrality is a bit more nuanced: How often does a vertex lies on the shortest path between any two given vertices?
  - I.e., you have to pass through _k_ to get from _i_ to _j_
  - A vertex with high betweenness -> mediating between others, ability to control information flow, e.g.
- Many different flavors of betweenness
  - `igraph` has two: original (vertex betweenness) and edge betweenness
  - `statnet` has many more
  - Borgatti, S. P., & Everett, M. G. (2006). A Graph-theoretic perspective on centrality. Social Networks, 28(4), 466–484.

#### Betweenness

```{r betweenness}
igraph::betweenness(g, directed = T)
### sna::betweenness(net, cmode = "directed")
```

#### Others, not discussed

- Many other nuanced measures of centrality
  - Eigenvector, closeness, Bonacich
  - Both packages have these measures
  
### Graph-level Indices

#### Properties of networks as a whole

- _Density_ - Total edges in a network : total possible number of edges

$$ Possible \ edges \ =n * (n-1) \ or \ \frac{1}{2} n * (n - 1) $$

- _Centralization_ - How much "inequality" is there in centrality scores in a network?
  - Measured relative to a theoretical maximum for inequality

```{r star network, echo = T, message = F, fig.asp=0.5}
make_star(10) %>% 
  ggraph() +
  geom_node_point() +
  geom_edge_link(alpha = 0.2) +
  theme_void()
```

#### `igraph` and `statnet` play a bit better here...

```{r graph-level}
### igraph
edge_density(g)

### statnet 
gden(net)
```

#### For centralization...

- `igraph` has individual functions with a prefix `centr_*`
- `statnet` has one function with many options

```{r centralization}
### igraph
centr_degree(g)

### statnet
### centralization(net, degree)
```

#### Subgraphs, communities

- `igraph` is vastly superior here -- has everything `statnet` has and more
- Many ways to determine cohesive communities within a network
  - Cliques, cores, plexes
  - Blockmodels, community detection

#### Community detection in `igraph`

- `cluster_*` prefix
- Computationally expensive
- Directedness of network often matters (with directed networks much more taxing)

```{r community}
fg <- cluster_fast_greedy(as.undirected(g))
fg

V(g)$mem <- fg$membership


```

#### Visualization of communities

```{r community map, message = F, echo = T, fig.width = 8, fig.height= 6}
ggraph(g) +
  geom_node_point(aes(color = as.factor(mem))) +
  geom_edge_link(alpha = 0.2) +
  theme_void() +
  guides(color=guide_legend(title="membership"))
```

### Visualization

#### So many options...

- Goals of network visualization: 
  - Create an easy to interpret and meaningful image of the network
  - Display attributes of vertices, edges
- Both `igraph` and `statnet` support visualiztions of networks
  - `igraph` (IMO) is produces better visualizations
  - Check out the help file -- many, many arguments and options, lots to tweak

#### Basics

- Three basics parameters for visualizing networks...
  - Layout - how are the vertices and edges arranged?
  - Vertices - color, size
  - Edges - color, size, transparency, arrows
- Invaluable resource: [http://kateto.net/network-visualization](http://kateto.net/network-visualization)

#### Layouts

- "Fixed" arrangements like circles, grids, trees
- Algorithmic layouts - treat networks as physical systems
  - Fruchterman-Reingold is a popular one
  - Very fiddly
- Choose based on what you are trying to communicate with the network
  - If you want "related" vertices to cluster, use an algorithmic approach

#### Layouts
- Hot tip: say the layout you want for multiple visualizations
  - `igraph` prefix: `layout_*`

```{r layouts with igraph}
l <- layout_with_fr(g)
head(l)
```

#### Multiple plots, same layout
  
```{r same layout, echo = T, fig.width = 8, fig.height = 5}
par(mfrow=c(1,3), mar=c(1, 1, 1, 1))
plot(g, layout = l, 
     vertex.color = V(g)$grade, vertex.label = NA, vertex.size = 5,
     edge.arrow.size = 0.5, main = "Grade")
plot(g, layout = l, vertex.color = as.factor(V(g)$race), 
     vertex.label = NA, vertex.size = 5,
     edge.arrow.size = 0.5, main = "Race")
plot(g, layout = l, 
     vertex.color = as.factor(V(g)$sex), vertex.label = NA, vertex.size = 5,
     edge.arrow.size = 0.5, main = "Sex")
```

#### Visualization in `igraph`

- `plot()` function
- Key arguments
  - `layout = ...`
  - Everything related to vertices uses the prefix `vertex.`
    - `vertex.size`, `vertex.color`
  - Everything related to edges uses the prefix `edges.`
    - `edge.color`, `edge.width`, `edge.arrow.size`

```{r igraph visualization, results=F}
par(mfrow=c(1,3), mar=c(1, 1, 1, 1))
plot(g, layout = l, 
     vertex.color = V(g)$grade, vertex.label = NA, vertex.size = 5,
     edge.arrow.size = 0.5, main = "Grade")
plot(g, layout = l, vertex.color = as.factor(V(g)$race), 
     vertex.label = NA, vertex.size = 5,
     edge.arrow.size = 0.5, main = "Race")
plot(g, layout = l, 
     vertex.color = as.factor(V(g)$sex), vertex.label = NA, vertex.size = 5,
     edge.arrow.size = 0.5, main = "Sex")
```

#### There isn't full overlap in parameters

```{r igraph vs statnet visualization, fig.width = 8, fig.height = 5}
par(mfrow=c(1,2), mar=c(1, 1, 1, 1))
plot(fg, g, layout = l, vertex.color = V(g)$grade, vertex.label = NA, vertex.size = 5, edge.arrow.size = 0.5, edge.color = "gray50", main = "Communities")
plot(net, vertex.col = (net %v% "grade"), edge.col = "gray50", usecurve = T, edge.curve = 0.01, main = "Curved edges")
```

#### A newcomer...`ggraph`

- `ggraph` is new visualization package built on the grammar of graphics idiom
  - Actually, there are several (`ggnet`, `ggnet2`, `ggnetwork`)
- Incorporates network visualiation into `ggplot2`
- `ggraph(graph, layout) + geom_node_* + geom_edge_*`
- Much easier, if you are familiar with `ggplot2` already

#### Visualization with `ggraph`

```{r ggraph}
library(ggraph)
ggraph(g, layout = "nicely") +  # works with igraph and statnet objects
  geom_node_point(aes(color = as.factor(grade))) +
  geom_edge_link(alpha = 0.2)
```

#### The power of `ggplot2`!

```{r ggraph facet}
library(ggraph)
ggraph(g, layout = "nicely") +  # works with igraph and statnet objects
  geom_node_point(aes(color = as.factor(grade))) +
  geom_edge_link(alpha = 0.2) +
  facet_nodes(~grade)
```

### Resources

- Basic network analysis
  - Hannenman & Riddle online textbook: http://faculty.ucr.edu/~hanneman/nettext/
- Visualization
  - Katya Ognyanova's guide: http://kateto.net/network-visualization
  - Thomas Lin Pedersen's guide to `ggraph`: https://www.data-imaginist.com/2017/ggraph-introduction-layouts/
- Statnet users group: http://www.statnet.org/statnet_users_group.shtml
- Stack Overflow's 'igraph' tag